\bf A note on the instability and degeneracy of deep learning models

### 2017-09-30 writing/paper-note/chapters/instab.Rmd
remarks don't link

For illustration, consider the iid Bernoulli model \@ref(eq:mod1) for $\boldsymbol X_N=(X_1,\ldots,X_N)$ with log-odds ratio parameter $\boldsymbol \theta_N = \log[ P_{\boldsymbol \theta_N}(X_1=1)/ P_{\boldsymbol \theta_N}(X_1=0)]\in\mathbb{R}$.  Remark \@ref(rmk1) (Section \@ref(criterion)) then gives the model instability measure \@ref(eq:Sun) directly as
$$
\frac{1}{N}\REP(\boldsymbol \theta_N ) = |\boldsymbol \theta_N|\quad\; \mbox{(iid Bernoulli model)},
$$
so that an unstable (or stable) model results for a divergent (or bounded) parameter sequence  $|\boldsymbol \theta_N|$. The above instability expression for the Bernoulli model follows as well from  the $N$-scaled LREP value \@ref(eq:UL) for a one-parameter exponential distribution, using that the sufficient statistic involved $g_N(\boldsymbol x)= \sum_{i=1}^N x_i$, $\boldsymbol x =(x_1,\ldots,x_n)\in\{0,1\}^N$, has maximum and mimimum values $U_N=N$ and $L_N=0$. In this case, @schweinberger2011instability has noted that the statistic is stable (i.e., bounded $(U_N-L_N)/N=1$) and the Bernoulli model is as well when, in particular, $\boldsymbol \theta_N=\bold \theta \in\mathbb{R}$ is fixed for $N \geq 1$.

@@

Proposition \@ref(prp:prop1) provides a sufficient condition for the stability of linear exponential models with fixed parameter dimension $k\geq 1$, whereby an S-stable model is guaranteed if the compounded magnitude of each combination of parameter $\theta_{i,N}$ and sufficient statistic value $(U_{i,N}-L_{i,N})$  is bounded by the sample size $N$, $i=1,\ldots,k$. This  supports the findings of @schweinberger2011instability, who showed degeneracy follows in such models under one type of violation of the condition \@ref(eq:prop1) in Proposition \@ref(prp:prop1) (namely, involving $k>1$ non-zero parameters  with $k-1$ statistics being $O(N)$ bounded while one statistic diverges in maximal size faster than the number $N$ of observations). To further illustrate the result in Proposition \@ref(prp:prop1), consider the multinomial distribution \@ref(eq:mod11) for $\boldsymbol X_N=(X_1,\ldots,X_N)$ having $k\geq 2$ categories $\{1,\ldots,k\}$ and $k$ parameters $\boldsymbol \theta_N = (\theta_{1,N},\ldots,\theta_{k,N})^\prime$. The variables are iid under this model so that Remark \@ref(rmk1) (Section \@ref(criterion) yields the corresponding $N$-scaled  log-ratio of extreme probabilities \@ref(eq:elpr) as
\begin{align*}
  \frac{1}{N}\REP(\boldsymbol \theta_N) &= \frac{\max_{1 \leq i \leq k} P_{\boldsymbol \theta_N}(X_1=i)}{\min_{1 \leq i \leq k} P_{\boldsymbol \theta_N}(X_1=i)}\\
  &= \max_{1 \leq i \leq k} \theta_{i,N} - \min_{1 \leq i \leq k} \theta_{i,N} \qquad \mbox{(iid multinomial model)}.
\end{align*}
Hence, a multinomial model sequence is unstable (or stable) depending on whether (or not) the maximal parameter difference $\max_{1 \leq i \leq k} \theta_{i,N} - \min_{1 \leq i \leq k} \theta_{i,N}$ diverges. Furthermore, using that each of the $k$ sufficient (count) statistics from the multinomial model \@ref(eq:mod11) satisfies $(U_{i,N}-L_{i,N})/N=1$, we  see that \@ref(eq:prop1) of Proposition \@ref(prp:prop1) becomes purely a parameter condition, $\sup_{N \geq 1}\max_{1 \leq i \leq k } |\theta_{i,N}| <\infty$, for ensuring that $\REP(\boldsymbol \theta_N)/N =\max_{1 \leq i \leq k} \theta_{i,N} - \min_{1 \leq i \leq k} \theta_{i,N}$ is bounded and stability follows for the multinomial distribution. Additionally, a stable multinomial sequence (i.e., bounded $\REP(\boldsymbol \theta_N)/N$) turns out to be nearly equivalent to \@ref(eq:prop1)  (e.g., these are the same if the smallest  parameter $\min_{1 \leq i \leq k } |\theta_{i,N}|$ remains bounded).

>>>>

For illustration, consider the iid Bernoulli model \@ref(eq:mod1) for $\boldsymbol X_N=(X_1,\ldots,X_N)$ with log-odds ratio parameter $\boldsymbol \theta_N = \log[ P_{\boldsymbol \theta_N}(X_1=1)/ P_{\boldsymbol \theta_N}(X_1=0)]\in\mathbb{R}$.  Remark 1 (Section \@ref(criterion)) then gives the model instability measure \@ref(eq:Sun) directly as
$$
\frac{1}{N}\REP(\boldsymbol \theta_N ) = |\boldsymbol \theta_N|\quad\; \mbox{(iid Bernoulli model)},
$$
so that an unstable (or stable) model results for a divergent (or bounded) parameter sequence  $|\boldsymbol \theta_N|$. The above instability expression for the Bernoulli model follows as well from  the $N$-scaled LREP value \@ref(eq:UL) for a one-parameter exponential distribution, using that the sufficient statistic involved $g_N(\boldsymbol x)= \sum_{i=1}^N x_i$, $\boldsymbol x =(x_1,\ldots,x_n)\in\{0,1\}^N$, has maximum and mimimum values $U_N=N$ and $L_N=0$. In this case, @schweinberger2011instability has noted that the statistic is stable (i.e., bounded $(U_N-L_N)/N=1$) and the Bernoulli model is as well when, in particular, $\boldsymbol \theta_N=\bold \theta \in\mathbb{R}$ is fixed for $N \geq 1$.

@@

Proposition \@ref(prp:prop1) provides a sufficient condition for the stability of linear exponential models with fixed parameter dimension $k\geq 1$, whereby an S-stable model is guaranteed if the compounded magnitude of each combination of parameter $\theta_{i,N}$ and sufficient statistic value $(U_{i,N}-L_{i,N})$  is bounded by the sample size $N$, $i=1,\ldots,k$. This  supports the findings of @schweinberger2011instability, who showed degeneracy follows in such models under one type of violation of the condition \@ref(eq:prop1) in Proposition \@ref(prp:prop1) (namely, involving $k>1$ non-zero parameters  with $k-1$ statistics being $O(N)$ bounded while one statistic diverges in maximal size faster than the number $N$ of observations). To further illustrate the result in Proposition \@ref(prp:prop1), consider the multinomial distribution \@ref(eq:mod11) for $\boldsymbol X_N=(X_1,\ldots,X_N)$ having $k\geq 2$ categories $\{1,\ldots,k\}$ and $k$ parameters $\boldsymbol \theta_N = (\theta_{1,N},\ldots,\theta_{k,N})^\prime$. The variables are iid under this model so that Remark 1 (Section \@ref(criterion) yields the corresponding $N$-scaled  log-ratio of extreme probabilities \@ref(eq:elpr) as
\begin{align*}
  \frac{1}{N}\REP(\boldsymbol \theta_N) &= \frac{\max_{1 \leq i \leq k} P_{\boldsymbol \theta_N}(X_1=i)}{\min_{1 \leq i \leq k} P_{\boldsymbol \theta_N}(X_1=i)}\\
  &= \max_{1 \leq i \leq k} \theta_{i,N} - \min_{1 \leq i \leq k} \theta_{i,N} \qquad \mbox{(iid multinomial model)}.
\end{align*}
Hence, a multinomial model sequence is unstable (or stable) depending on whether (or not) the maximal parameter difference $\max_{1 \leq i \leq k} \theta_{i,N} - \min_{1 \leq i \leq k} \theta_{i,N}$ diverges. Furthermore, using that each of the $k$ sufficient (count) statistics from the multinomial model \@ref(eq:mod11) satisfies $(U_{i,N}-L_{i,N})/N=1$, we  see that \@ref(eq:prop1) of Proposition \@ref(prp:prop1) becomes purely a parameter condition, $\sup_{N \geq 1}\max_{1 \leq i \leq k } |\theta_{i,N}| <\infty$, for ensuring that $\REP(\boldsymbol \theta_N)/N =\max_{1 \leq i \leq k} \theta_{i,N} - \min_{1 \leq i \leq k} \theta_{i,N}$ is bounded and stability follows for the multinomial distribution. Additionally, a stable multinomial sequence (i.e., bounded $\REP(\boldsymbol \theta_N)/N$) turns out to be nearly equivalent to \@ref(eq:prop1)  (e.g., these are the same if the smallest  parameter $\min_{1 \leq i \leq k } |\theta_{i,N}|$ remains bounded).

### 2017-09-30 writing/paper-note/docs/note.tex
remarks don't link

For illustration, consider the iid Bernoulli model \eqref{eq:mod1} for
\(\boldsymbol X_N=(X_1,\ldots,X_N)\) with log-odds ratio parameter
\(\boldsymbol \theta_N = \log[ P_{\boldsymbol \theta_N}(X_1=1)/ P_{\boldsymbol \theta_N}(X_1=0)]\in\mathbb{R}\).
Remark \ref{rmk1} (Section \ref{criterion}) then gives the model
instability measure \eqref{eq:Sun} directly as \[
\frac{1}{N}\REP(\boldsymbol \theta_N ) = |\boldsymbol \theta_N|\quad\; \mbox{(iid Bernoulli model)},
\] so that an unstable (or stable) model results for a divergent (or
bounded) parameter sequence \(|\boldsymbol \theta_N|\). The above
instability expression for the Bernoulli model follows as well from the
\(N\)-scaled LREP value \eqref{eq:UL} for a one-parameter exponential
distribution, using that the sufficient statistic involved
\(g_N(\boldsymbol x)= \sum_{i=1}^N x_i\),
\(\boldsymbol x =(x_1,\ldots,x_n)\in\{0,1\}^N\), has maximum and mimimum
values \(U_N=N\) and \(L_N=0\). In this case, Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) has noted
that the statistic is stable (i.e., bounded \((U_N-L_N)/N=1\)) and the
Bernoulli model is as well when, in particular,
\(\boldsymbol \theta_N=\bold \theta \in\mathbb{R}\) is fixed for
\(N \geq 1\).

@@

Proposition \ref{prp:prop1} provides a sufficient condition for the
stability of linear exponential models with fixed parameter dimension
\(k\geq 1\), whereby an S-stable model is guaranteed if the compounded
magnitude of each combination of parameter \(\theta_{i,N}\) and
sufficient statistic value \((U_{i,N}-L_{i,N})\) is bounded by the
sample size \(N\), \(i=1,\ldots,k\). This supports the findings of
Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}), who showed
degeneracy follows in such models under one type of violation of the
condition \eqref{eq:prop1} in Proposition \ref{prp:prop1} (namely,
involving \(k>1\) non-zero parameters with \(k-1\) statistics being
\(O(N)\) bounded while one statistic diverges in maximal size faster
than the number \(N\) of observations). To further illustrate the result
in Proposition \ref{prp:prop1}, consider the multinomial distribution
\eqref{eq:mod11} for \(\boldsymbol X_N=(X_1,\ldots,X_N)\) having
\(k\geq 2\) categories \(\{1,\ldots,k\}\) and \(k\) parameters
\(\boldsymbol \theta_N = (\theta_{1,N},\ldots,\theta_{k,N})^\prime\).
The variables are iid under this model so that Remark \ref{rmk1}
(Section \ref{criterion} yields the corresponding \(N\)-scaled log-ratio
of extreme probabilities \eqref{eq:elpr} as
\begin{align*}
  \frac{1}{N}\REP(\boldsymbol \theta_N) &= \frac{\max_{1 \leq i \leq k} P_{\boldsymbol \theta_N}(X_1=i)}{\min_{1 \leq i \leq k} P_{\boldsymbol \theta_N}(X_1=i)}\\
  &= \max_{1 \leq i \leq k} \theta_{i,N} - \min_{1 \leq i \leq k} \theta_{i,N} \qquad \mbox{(iid multinomial model)}.
\end{align*}
Hence, a multinomial model sequence is unstable (or stable) depending on
whether (or not) the maximal parameter difference
\(\max_{1 \leq i \leq k} \theta_{i,N} - \min_{1 \leq i \leq k} \theta_{i,N}\)
diverges. Furthermore, using that each of the \(k\) sufficient (count)
statistics from the multinomial model \eqref{eq:mod11} satisfies
\((U_{i,N}-L_{i,N})/N=1\), we see that \eqref{eq:prop1} of Proposition
\ref{prp:prop1} becomes purely a parameter condition,
\(\sup_{N \geq 1}\max_{1 \leq i \leq k } |\theta_{i,N}| <\infty\), for
ensuring that
\(\REP(\boldsymbol \theta_N)/N =\max_{1 \leq i \leq k} \theta_{i,N} - \min_{1 \leq i \leq k} \theta_{i,N}\)
is bounded and stability follows for the multinomial distribution.
Additionally, a stable multinomial sequence (i.e., bounded
\(\REP(\boldsymbol \theta_N)/N\)) turns out to be nearly equivalent to
\eqref{eq:prop1} (e.g., these are the same if the smallest parameter
\(\min_{1 \leq i \leq k } |\theta_{i,N}|\) remains bounded).

>>>>

For illustration, consider the iid Bernoulli model \eqref{eq:mod1} for
\(\boldsymbol X_N=(X_1,\ldots,X_N)\) with log-odds ratio parameter
\(\boldsymbol \theta_N = \log[ P_{\boldsymbol \theta_N}(X_1=1)/ P_{\boldsymbol \theta_N}(X_1=0)]\in\mathbb{R}\).
Remark 1 (Section \ref{criterion}) then gives the model instability
measure \eqref{eq:Sun} directly as \[
\frac{1}{N}\REP(\boldsymbol \theta_N ) = |\boldsymbol \theta_N|\quad\; \mbox{(iid Bernoulli model)},
\] so that an unstable (or stable) model results for a divergent (or
bounded) parameter sequence \(|\boldsymbol \theta_N|\). The above
instability expression for the Bernoulli model follows as well from the
\(N\)-scaled LREP value \eqref{eq:UL} for a one-parameter exponential
distribution, using that the sufficient statistic involved
\(g_N(\boldsymbol x)= \sum_{i=1}^N x_i\),
\(\boldsymbol x =(x_1,\ldots,x_n)\in\{0,1\}^N\), has maximum and mimimum
values \(U_N=N\) and \(L_N=0\). In this case, Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) has noted
that the statistic is stable (i.e., bounded \((U_N-L_N)/N=1\)) and the
Bernoulli model is as well when, in particular,
\(\boldsymbol \theta_N=\bold \theta \in\mathbb{R}\) is fixed for
\(N \geq 1\).

@@

Proposition \ref{prp:prop1} provides a sufficient condition for the
stability of linear exponential models with fixed parameter dimension
\(k\geq 1\), whereby an S-stable model is guaranteed if the compounded
magnitude of each combination of parameter \(\theta_{i,N}\) and
sufficient statistic value \((U_{i,N}-L_{i,N})\) is bounded by the
sample size \(N\), \(i=1,\ldots,k\). This supports the findings of
Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}), who showed
degeneracy follows in such models under one type of violation of the
condition \eqref{eq:prop1} in Proposition \ref{prp:prop1} (namely,
involving \(k>1\) non-zero parameters with \(k-1\) statistics being
\(O(N)\) bounded while one statistic diverges in maximal size faster
than the number \(N\) of observations). To further illustrate the result
in Proposition \ref{prp:prop1}, consider the multinomial distribution
\eqref{eq:mod11} for \(\boldsymbol X_N=(X_1,\ldots,X_N)\) having
\(k\geq 2\) categories \(\{1,\ldots,k\}\) and \(k\) parameters
\(\boldsymbol \theta_N = (\theta_{1,N},\ldots,\theta_{k,N})^\prime\).
The variables are iid under this model so that Remark 1 (Section
\ref{criterion} yields the corresponding \(N\)-scaled log-ratio of
extreme probabilities \eqref{eq:elpr} as
\begin{align*}
  \frac{1}{N}\REP(\boldsymbol \theta_N) &= \frac{\max_{1 \leq i \leq k} P_{\boldsymbol \theta_N}(X_1=i)}{\min_{1 \leq i \leq k} P_{\boldsymbol \theta_N}(X_1=i)}\\
  &= \max_{1 \leq i \leq k} \theta_{i,N} - \min_{1 \leq i \leq k} \theta_{i,N} \qquad \mbox{(iid multinomial model)}.
\end{align*}
Hence, a multinomial model sequence is unstable (or stable) depending on
whether (or not) the maximal parameter difference
\(\max_{1 \leq i \leq k} \theta_{i,N} - \min_{1 \leq i \leq k} \theta_{i,N}\)
diverges. Furthermore, using that each of the \(k\) sufficient (count)
statistics from the multinomial model \eqref{eq:mod11} satisfies
\((U_{i,N}-L_{i,N})/N=1\), we see that \eqref{eq:prop1} of Proposition
\ref{prp:prop1} becomes purely a parameter condition,
\(\sup_{N \geq 1}\max_{1 \leq i \leq k } |\theta_{i,N}| <\infty\), for
ensuring that
\(\REP(\boldsymbol \theta_N)/N =\max_{1 \leq i \leq k} \theta_{i,N} - \min_{1 \leq i \leq k} \theta_{i,N}\)
is bounded and stability follows for the multinomial distribution.
Additionally, a stable multinomial sequence (i.e., bounded
\(\REP(\boldsymbol \theta_N)/N\)) turns out to be nearly equivalent to
\eqref{eq:prop1} (e.g., these are the same if the smallest parameter
\(\min_{1 \leq i \leq k } |\theta_{i,N}|\) remains bounded).
